---
output:
  md_document:
    variant: markdown_github
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
stopifnot(require(knitr))
knitr::opts_chunk$set(
  echo = TRUE, 
  eval = TRUE,
  fig.asp = 0.8,
  fig.fullwidth=TRUE,
#  out.width = "60%",
  fig.align = "center")
library(rstan)
library(sf)
library(spdep)
library(INLA)
library(ggplot2)
rstan_options(auto_write = TRUE)
#model = stan_model("BYM2.stan")
source("https://raw.githubusercontent.com/ConnorDonegan/geostan/main/R/convenience-functions.R")
source("https://raw.githubusercontent.com/ConnorDonegan/helpful/master/get-shp.R")
c_scale <- function(C) { # create scale factor for BYM2 model
    # C is N by N connectivity matrix
    N <- nrow(C)
    adj.matrix <- Matrix::Matrix(C, sparse = TRUE)
    Q =  Diagonal(N, rowSums(adj.matrix)) - adj.matrix
    Q_pert = Q + Diagonal(N) * max(diag(Q)) * sqrt(.Machine$double.eps)
    Q_inv = INLA::inla.qinv(Q_pert, constr=list(A = matrix(1,1,N),e=0))
    scale_factor <- exp(mean(log(diag(Q_inv))))
    return (scale_factor)
}
rstan_options(auto_write = TRUE)
```

## Fitting IAR/BYM2 models in Stan with disconnected graph structures

This repo contains R and Stan code to fit spatial models using intrinsic autoregressive priors, with or without a fully connected graph structure (i.e. islands and multiple disconnected regions are accepted). It builds on earlier work by [Mitzi Morris](https://github.com/stan-dev/example-models/blob/master/knitr/car-iar-poisson/nb_data_funs.R) and [Adam Howes](https://athowes.github.io/2020/11/10/fast-disconnected-car/) with some slight differences. The method here avoids any shuffling of observations in R; and here the BYM2 model (which does partial pooling across neighboring observations using parameter $\phi$, and across all observations using $\theta$) $\phi_i$ gets zeroed out for any islands or other observations with zero neighbors. 

For an introduction to these models see Morris' [Stan case study](https://mc-stan.org/users/documentation/case-studies/icar_stan.html).

### Stan code

The intrinsic autoregressive prior is programmed following Morris et al. 2020:

```{Rcpp eval = FALSE}
functions {
/**
  * intrinsic autoregressive prior (Morris et al 2019)
  * @return lpdf of IAR prior minus any constant terms
  */
  real icar_normal_lpdf(vector phi, int N, int[] node1, int[] node2) {
    return -0.5 * dot_self(phi[node1] - phi[node2]) +
      normal_lpdf(sum(phi) | 0, 0.001 * N);
  }
}
```

It can be broken into two parts: first the pairwise difference formula `-0.5 * dot_self(phi[node1] - phi[node2])` which places low (prior) probability on large differences between neighbors, and then the sum-to-zero constraint `sum(phi) ~ normal(0, 0.001*N)`.
 
It needs to be applied independently to each group of connected observations and islands may need some kind of special treatment. If there were no partial-pooling across all observations (i.e. there were no $\theta$, just $\phi$) we could follow what others have done and set the prior distributions for the islands to $\phi_i \sim N(0, 1)$. But with the BYM2 that will end up duplicating $\theta_i$. Instead, we'll zero out $\phi_i$ for any islands.

First we'll take Adam Howe's observation that the following code chunk already drops out the islands, so it can remain as is:

```{Rcpp eval = FALSE}
target += -0.5 * dot_self(phi[node1] - phi[node2]);
```
Only connected nodes are in those indices. 

Now we just need to add the sum-to-zero constraint, which again happens to take care of our island issue naturally. The data `group_idx` allows us to extract index observations by group; and we pull those indices out using Stan's `segment` function. If the `group_size` is one (an island), this gives it the strong prior distribution we need to zero it out (and we might add an extra zero to the scale, `0.0001`):
```{Rcpp eval = FALSE}
  int pos;
  pos = 1;
  for (j in 1:k) {
    sum(phi[segment(group_idx, pos, group_size[j])]) ~ normal(0, 0.001 * group_size[j]);
    pos = pos + group_size[j];
  }
```

Then the full model in `BYM2.stan` is below, which will work with or without a fully connected graph structure and doesn't require any reshuffling of observations. It requires needs the following input:

- `n` number of observations
- `k` number of groups of connected observations (i.e. a connected graph with one disconnected island has `k=1`)
- `group_size` an integer array with the number of observations per group
- `n_edges` total number of edges (sum of all `edge_size`)
- `node1`, `node2` same as in Morris' code, these contain the indices for each pair of connected nodes
- `group_idx` integer array (size `n`), see below. Allows us to extract observations by their group membership.
- `scale_factor` vector of scale factors, one per group---but assigned to every observations according to group membership
- `prior_only` Binary indicator, skip the likelihood and just compute the prior or sample from posterior distribution given data
- `y` outcome data (integer array) (ignored if `prior_only=1`)
- `offset` offset term (gets log transformed internally) (ignored if `prior_only=1`)

```{stan output.var="model"}
// BYM2 model with possibly disconnected group structure and islands
data {
  int n;    // no. observations
  int k;    // no. groups
  int group_size[k]; // group sizes---to extract data by group
  int<lower=0> n_edges; // total number of edges
  int<lower=0> node1[n_edges];
  int<lower=0> node2[n_edges];
  int group_idx[n]; // e.g. phi[group_idx[1:group_size[1]]] contains group 1 outcomes
  vector[n] scale_factor; // will repeat same value for all observations by group;
  int<lower=0, upper=1> prior_only;
  int y[n];
  vector[n] offset;
}

transformed data {
  vector[n] log_off = log(offset);
}

parameters {
  real alpha;
  real<lower=0> sigma_re;
  vector[n] phi;
  vector[n] theta;
  real logit_rho;
}

transformed parameters {
  vector[n] convolved_re;
  real<lower=0, upper=1> rho;
  rho = inv_logit(logit_rho);
  convolved_re = sqrt(rho * inv(scale_factor)) .* phi + sqrt(1 - rho) * theta;      
}

model {
  int pos;
    pos = 1;
    for (j in 1:k) {
      sum(phi[segment(group_idx, pos, group_size[j])]) ~ normal(0, 0.001 * group_size[j]);
      pos = pos + group_size[j];
    }
   target += -0.5 * dot_self(phi[node1] - phi[node2]);
   sigma_re ~ std_normal();
   theta ~ std_normal();
   logit_rho ~ std_normal();
   alpha ~ std_normal();
   if (!prior_only) y ~ poisson_log(log_off + alpha + convolved_re * sigma_re);
}

```

### Demonstration

This demonstrates using all US states (including Puerto Rico and D.C.). 
```{r eval = FALSE}
library(rstan)
rstan_options(auto_write = TRUE)
library(sf)
library(spdep)
library(INLA)
library(ggplot2)
model = stan_model("BYM2.stan")
# load geostan::edges, geostan::shape2mat 
source("https://raw.githubusercontent.com/ConnorDonegan/geostan/main/R/convenience-functions.R")
# load get.shp function to download shapefiles
source("https://raw.githubusercontent.com/ConnorDonegan/helpful/master/get-shp.R")
c_scale <- function(C) { # create scale factor for BYM2 model
    # C is N by N connectivity matrix
    N <- nrow(C)
    adj.matrix <- Matrix::Matrix(C, sparse = TRUE)
    Q =  Diagonal(N, rowSums(adj.matrix)) - adj.matrix
    Q_pert = Q + Diagonal(N) * max(diag(Q)) * sqrt(.Machine$double.eps)
    Q_inv = INLA::inla.qinv(Q_pert, constr=list(A = matrix(1,1,N),e=0))
    scale_factor <- exp(mean(log(diag(Q_inv))))
    return (scale_factor)
}
```

Download the shapefile from the Census Bureau and load it as an `sf` (simple features) object:

```{r}
## get a shapefil
url <- "https://www2.census.gov/geo/tiger/GENZ2019/shp/cb_2019_us_state_20m.zip"
#get.shp(url, "states")
states <- st_read("states")
ggplot(states) +
  geom_sf() +
  theme_void()
```

Much of the following operations could be combined together into a single loop but is shown here with results:

```{r}
## create a list of neighbors based on contiguity (shared borders)
nb <- spdep::poly2nb(states)
nb2 <- spdep::n.comp.nb(nb)
print(nb2)

## make group_idx out of nb2
n <- nrow(states)
k = nb2$nc
group_idx = NULL
for (j in 1:k) {
    group_idx <- c(group_idx, which(nb2$comp.id == j))
}
print(group_idx)

## group sizes (number of observations per group)
sizes <- NULL
for (j in 1:k) {
    sizes <- c(sizes, sum(nb2$comp.id == j))
}
print(sizes)

## make edge list 
C <- shape2mat(states, style = "B")
E <- edges(C)
print(head(E))

## make scale_factors
scale_factor <- vector(mode = "numeric", length = n)
for (j in 1:k) {
    g.idx <- which(nb2$comp.id == j) 
    if (length(g.idx) == 1) {
        scale_factor[g.idx] <- 1 
        next
    }    
    Cg <- C[g.idx, g.idx] 
    scale.j <- c_scale(Cg)
    scale_factor[g.idx] <- scale.j
}
print(scale_factor)
```

And put it in a list and feed to Stan:
```{r}
dl <- list(
    n = n,
    k = k,
    group_size = array(sizes, dim = k),
    n_edges = nrow(E),
    node1 = E$node1,
    node2 = E$node2,
    group_idx = group_idx,
    scale_factor = scale_factor,
    prior_only = TRUE,
    y = rep(1, n), # just a placeholder
    offset = rep(1, n) # placeholder
)

fit = sampling(model,
               data = dl,
               chains = 4,
               cores = 4,
               control = list(max_treedepth = 13)
               )
```

We can see that three of the $\phi_i$ are zero (the disconnected areas):
```{r fig.fullwidth=TRUE}
plot(fit, pars = "phi")
```

And here's the (mean) prior degree of spatial autocorrelation in $\phi$:
```{r}
W <- shape2mat(states, "W")
phi <- as.matrix(fit, pars = "phi")
phi  <- apply(phi, 2, mean)
moran_plot(phi, W)
```



